---
title: "The zorgmijding package"
author: "Premysl Velek"
date: "12 february 2023"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This package takes calculates and plots GP contact rate for
cardiovascular complaints before and during the COVID-19 pandemic. It
was developed specifically for the ZonMw 'Zorgmijding' project and works
ONLY with data exported form GP registration databases managed by five
University Medical Centres in the Netherlands (EMC Rotterdam, UMC
Utrecht, UMC Amsterdam, UMC Mastricht and UMC Gronningen).

# Using the package

There are four different functions in the package, covering the entire
pipeline from cleaning the raw data to plotting the final results. Those
function are:

1.  `clean_data` takes raw data in csv format and cleans it for further
    analysis
2.  `denominators` calculates the population size (ie. the number of
    patients in the database)
3.  `n_visits` calculates the consultation rates in the pre-pandemic
    period and in 2020
4.  `plot_visits` plots the consultation rate

The functions have to be run in this specific order, as the output in
one is used as an input to the subsequent function. However, it is
entirely possible (and recommended) to re-use the output produced by the
functions and all functions (except `plot_visit`) store their output on
your computer. For example, you can run the `n_visit` using the output
form the `clean` and `denominators` that is stored on your computer.

To use the package, first install it and load it onto R:

```{r load zorgmijding, eval = FALSE, include = TRUE}
if(!require(devtools)) install.packages("devtools")
devtools::install_local("path/to/folder/zorgmijding_0.0.0.9000.tar.gz", dependencies = TRUE)
library(zorgmijding)
```

## Cleaning

The `clean_data` function takes raw data and format
and cleans it for further analysis This is the fist function in this
pipeline, it should be run first. It makes sure the variables are
consistently named, are in the same order and have the right format. It
also check there are no missing or nonsensical data.

After running the function, you will be asked to provide two raw data
files in the csv format: the one with GP contacts and one with all
patients. The cleaned data are automatically saved in the 'results'
folder created as a sub-folder in the folder in which the raw data are
located.

The function has two arguments. The first argument -umc - specifies the UMC that 
provided the data; it can take the following values: "utrecht", "maastricht",
"amsterdam", "groningen", "rotterdam". The second argument - clean_types - indicates
whether the list of eligible contact types should be used when cleaning the data. 
It only makes sense to use for data that does not come from Rotterdam, so when it's 
set to TRUE and the umc is set to 'rotterdam', the it changes automatically to FALSE 
with a warning message printed out. The default value is FALSE.  

```{r example clean, eval = FALSE, include = TRUE}
clean_data(umc = "rotterdam")
```

## Calculating sample size

The `denominators` function takes the cleaned patient data - saved by
the `clean_data` function - to calculate the population size for each
period - month, week and day - from 2016 to 2020. It does it for all
patients, but also by age groups, by sex and by age and sex. The results
are automatically saved in the results' folder created as a subfolder in
the folder in which the raw data are located.

After runinng the function, you will be asked to provide the cleaned
patients data (in .RData format).

The function has no arguments.

**NOTE:** Running the function may take a while (up to several hours),
depending on the size of your database.

```{r example denominators, eval = FALSE, include = TRUE}
denominators()
```

## Counting GP contacts

The `n_visits` function takes the cleaned visit data (output of the
`clean_data` function) and the denominators data (output of the
`denominators` function) and counts the number of GP contacts per period
(month, week, day) and per 100,000 patients. The output is stored as
separate data frames for daily, weekly and monthly contact rates,
stratified by sex, age, sex and age and as a total number. You can also
specify whether you want to see all data, or only data for patients 40
and older. In total 18 different data frames are saved.

The function has one argument averages an indicator whether a
prepandemic averages should be computed. If set to TRUE (the default),
then the output is a dataset in which there are two numbers for each
period (month, week, day) in a year: one is the pre-pandemic average
(2017-2019) weighted by the population size, and the other is the
observed values in 2020. If set to FALSE, then data for the entire
period (2016-2020) is saved. If set to FALSE, the function `plot_visits`
will not work.


```{r example n_visits n 1, eval = FALSE, include = TRUE}
n_visits(averages = "TRUE")
```

The function returns the path to the RData file with the resulting data frames. 
You can therefore use it as an input to the plotting function (see below). 
(If the path is not defined, the function will ask the user to select the relevant 
file.)

```{r example n_visits n 2, eval = FALSE, include = TRUE}
path_to_res <- n_visits(averages = "TRUE")

plot_visits("Weekly rate of GP contacts for cardiovascular complaints by sex",
  stratum = "sex", periodicity = "w", segment = "full",
  show_40plus = "TRUE", 
  path = path_to_res
)
```


## Plotting

The function `plot_visits` plots the visit counts data, the output of
the `n_visit` function. It only works when the the pre-pandemic averages
are computed, ie. when the argument "averages" in the `n_visits`
function was set to "TRUE"

There are several arguments that need to be specified: \* Title of the
plot \* The stratification variable, possible values are 'sex', 'age',
'sex_age', 'total' (the default). Plot by sex will be shown on one
panel, plots by age will be faceted across multiple panels \* The period
of counts, can either be daily, weekly or monthly. The values are 'd':
day, 'w': week or 'm': month (the default). If plotting daily rate, it
is highly recommended to \* Time period to be plotted. By default, the
data are plotted for the entire year, with the pre-pandemic averages
plotted against the 2020 data. If plotting daily data, it is recommended
to use shorter period. The period is specified by two dates: start and
end date of the period to be plotted. \* Whether you want to plot data
for the entire population or only for patients of 40 years and older \*
Path to the data to be plotted. Optional, if not provided you will be
prompted to select the data via GUI after running the function.

```{r example plot_visits, eval = FALSE, include = TRUE}
plot_visits("Weekly rate of GP contacts for cardiovascular complaints by sex",
  stratum = "sex", periodicity = "w", segment = "full",
  show_40plus = "TRUE"
)

plot_visits("Daily rate of GP contacts for cardiovascular complaints by age",
  stratum = "age", periodicity = "d",
  segment = c("2020-02-15", "2020-04-31"),
  show_40plus = "TRUE"
)
```
